Placing the spotted T Tauri star LkCa 4 on an HR diagram

### 2017-01-03 document/submission/ms.tex
remove the bf comments from GJH

Ages and masses of young stars are often estimated by comparing their observed luminosities and effective temperatures to pre-main sequence stellar evolutionary model tracks, but nonstandard physics like magnetic fields and starspots complicate this process. In order to understand their influence, we study in detail the heavily-spotted weak-lined \emph{T-Tauri} star LkCa 4 by searching for spectral signatures of radiation originating from the starspot or starspot groups.  We introduce a new methodology for constraining both the starspot filling factor and the typical spot temperature by fitting two-temperature stellar atmosphere models constructed from \PHOENIX\ synthetic spectra to a high-resolution near-IR IGRINS spectrum.  Clearly discernable spectral features arise from both a hot photospheric component $\teffa\sim4100$ K and to a cool component $\teffb\sim2700-3000$ K, {\bf with a best-fit filling factor of $\sim80$\%  for the cool photosphere at the time of the observation.} {\bf These results are supported by comparisons of this mix of hot and cool photospheric temperatures by analyses of the spectral energy distribution, rotational modulation of colors and of TiO band strengths, and low-resolution optical/near-IR spectroscopy.  LkCa 4 has either a warm photosphere mostly covered in cool spots or is a very cool star with a small hot spot.}  The revised effective temperature and luminosity make LkCa 4 appear much younger and lower mass when compared to previous estimates and naively using the same model tracks.  Biases from starspots likely afflict most {\bf fully convective} young stars and contribute to uncertainties in open cluster ages inferred from low-mass stars and in large age spreads of those clusters. We also discuss the detailed influence of starspots on the IGRINS spectrum, which in some regions acts as a relatively featureless ``veiling'' continuum owing to the large rotational broadening ($\vsini\sim28$ km~s$^{-1}$) and heavy line-blanketing in cool star spectra, {\bf and anti-correlations expected in the velocity jitter of the warm and cool components.  I put this in cause I think it's interesting, but delete if you do not want to highlight this in the abstract.}

@@

{\bf The causes of age uncertainties and of large luminosity spreads in young clusters are controversial \citep[see reviews by][]{preibisch12,soderblom14}.  Absolute ages are usually inferred from comparisons between observed stellar properties and pre-main sequence evolutionary models, which are sensitive to stellar birthlines and the input physics that controls the contraction and internal heating.  Within clusters, any real differences in ages between stars may be masked by observational uncertainties in temperature and luminosity measurements \citep[e.g.][]{hartmann01,slesnick08}.}

@@

In this work, we measure the starspot properties of LkCa 4 with four complementary techniques.  Section \ref{sec:obs} describes the full spectroscopic and photometric dataset and 
quantifies the optical variability of LkCa 4 over the last 31 years based on all available photometric monitoring; a collection of spectral observations at various phases of variability is introduced.  Section 3 and Appendix \ref{methods-details} describe a spectral inference technique aimed at generalizing line depth ratio analysis, which is applied to a high resolution $H$ and $K$ near-IR spectrum of LkCa 4 in Section \ref{sec:two_tempIGRINS}.  Section \ref{sec:GJHsection4} shows results from SED fitting, polychromatic photometric monitoring, and optical TiO line-depth ratio fitting.  All lines of evidence are ultimately combined to build a consistent picture of the spectral and temporal evolution of LkCa 4.  {\bf Section 5 describes how the two-temperature fit affects the placement of LkCa 4 on the HR diagram and what can be understood about pre-main sequence stellar evolution from this exemplar.  Finally, Section 6 describes how our results should be interpreted, the implications for analyses of spots in high-resolution spectra, and methodological limitations.}

@@

Starspots possess a spectrum distinct from the ambient photosphere.  Approaches like ZDI and line-bisector analysis \citep[\emph{e.g.}][]{prato08, donati14} target the modulation of stellar photospheric lines as starspots enter and exit the observable stellar disk.  In contrast, the technique described in this Section measures the spectrum that emerges from {\bf both the hot and the cool surfaces of the photosphere.}

@@

{\bf reordered to place main result first}

@@

Overplotting forward-modeled spectra with the observed IGRINS spectra yielded insights on why some spectral orders perform better than others in assessing stellar properties.  Orders with extremely poor telluric correction residuals, large spectral line outliers, and uncorrected H line residuals from A0V standard division, are all excluded from our final stellar parameter compilation.  Several orders in $K-$band were rejected due to poor spectral fits influenced by deep metal lines.  For example, orders 79 and 81 {\bf I'm not really sure what {\it overfit} means} overfit Ca~I and Na~I lines and orders $74-76$ overfit the CO lines, which are known for their gravity sensitivity \citep{rayner09}.  These and other metal lines can also be biased by Zeeman broadening, which could explain poorer fit quality with wavelength \citep{johnskrull07,deen13}.  Additionally, some orders without these conspicuous faults, but simply possessing mediocre fits, or relatively uninformative fits, were discarded.  The remaining orders are relatively devoid of spectral line outliers and include the most information rich portion of the spectrum.

@@

%\caption{Top:  The low-resolution optical/near-IR spectrum of LkCa 4 obtained from Palomar/DBSP and APO/Triplespec on 30 December 2008 (black), compared to a synthetic spectrum of a two temperature photosphere (purple).  The inset shows that the 2750 K (red, 80\% fill factor) and 4100 K (blue, 20\% fill factor) components contribute equally to the near-IR spectrum, but the 4100 K component dominates the blue emission.  The synthetic spectrum is reddened by $A_V=0.3$ mag and scaled to the observed $J$-band spectrum. Bottom:  The low-resolution optical (left) and near-IR (right) spectrum of LkCa 4, compared with a 3900 K photosphere (blue), a 3500 K photosphere (red), and the two temperature photosphere (pink) that best fits the IGRINS spectrum.  The synthetic spectra are scaled separately to the optical spectrum at 0.75 $\mu$m and to the near-IR spectrum at 1.5 $\mu$m.  Warm photospheres accurately reproduce molecular bands at $0.7$ $\mu$m but fail to fit the spectral features at longer wavelengths.  Cooler photospheres predict molecular bands at $<0.7$ $\mu$m that are much deeper than observed.  The two temperature photosphere accurately fits spectral features in both optical and near-IR wavelengths.}

@@

The synthetic models match the full spectrum reasonably well.  When scaled separately to the red-optical and near-IR spectra, the synthetic models match even better (bottom panels of Figure \ref{fig:lores}), while single temperature photospheres fail to reproduce large spectral features.   The $3900$ K component fits the TiO bands at $<7400$ \AA\ but fails to reproduce molecular bands at longer wavelengths.  The $3500$ K component suffers from the opposite problem, yielding molecular bands at short wavelengths that are too deep.  The two temperaturephotosphere with the parameters from the best-fit calculated in \S 3 reproduces the TiO bands, the hump in the $H$-band, and the jump in flux at the long-wavelength end of the $J$-band.  {\bf This latter feature is caused by H$_2$O opacity in H$_2$O bands and is only seen when the fraction of emission from the cool component is high (Figure \ref{fig:h2ojump}).}

@@

The TiO bands trace the cool component.  When the hotter component has a larger filling fraction, the TiO bands are veiled by the hotter component and should therefore be shallower. As expected, 
the TiO indices all correlate strongly with the optical brightness.  {\bf The TiO-7140 band depths are accurately reproduced by varying the two-temperature fit with rotation.  Other bands have small offsets, likely caused by either contributions from intermediate temperatures or by small errors in the synthetic spectra.}  In the spectral type scheme derived by \citet{herczeg14}, the change in TiO-7140 index corresponds to a range from M1.5 (3640 K) to M2.2 (3530 K).

@@

An accurate extinction estimate from the full SED is challenging because of a strong dependence on the  distribution of temperatures in the photosphere, which in our work is simplified to a two temperature fit.  However, isolating specific colors constrains the possible range of extinctions.  {\bf Based on the following analysis, an $A_V=0.3$ mag is adopted throughout this paper, although some uncertainty is introduced by confusion in the choice of template colors.}

@@

Our results tentatively support the lower extinctions obtained at optical wavelengths, although the template colors used to calculate extinction are uncertain and will need revision.  Differences between main sequence and pre-main sequence temperature scales and colors in the \citet{pecaut13} tables are introduced by a combination of spots, in addition to any color differences caused by gravity differences.
Uncertainty in colors may be solved with self-consistent comparisons to young stars with similar spectral types, if spot coverage is also similar.  Discrepancies in colors, spectral type/effective temperatures, and extinctions for LkCa 4, and perhaps more broadly for TTSs, are dominated by ignoring spots and not by spectral type-effective temperature scales.  {\bf Fundamental problems in molecular opacities and photospheric models are likely important secondary affects.}

@@

{\bf just a note, I do really like this subsection, it's very simple.}

@@

The Pleiades serves as an older {\bf \citep[$\sim125$ Myr][]{stauffer98}} sample of young stars, many of which have starspots that cover large fractions of the stellar surface.  For example, anomalous colors of K dwarfs in the Pleiades can be explained by $\sim$50\% fill factors of spots \citep{stauffer03}, while LAMOST spectra of the TiO band indicates that $\sim 30$\% spot coverages may be common for 3500--5000 K stars \citep{fang2016}.  Recent photometric monitoring of the Pleiades indicates that the anomalous colors scale with rotation rate, strengthing the case for magnetic-field induced starspots as the cause of the anomaly \citep{covey16}.  However, the \citet{covey16} Pleiades rotation rates do not scale with the $\Delta V$ photometric amplitude at the precision of the PTF data, suggesting that while the overall starspot coverage increases with rotation rate, the longitudinally asymmetric component does not.  LkCa 4 represents an example of such an architecture, with more than  50\% of the visible stellar surface covered in spots at all observational phases. The longitudinally symmetric component evades photometric modulation detection, but contributes to otherwise anomalous color offsets.

Spots are also likely important in estimating the properties of generic low-mass young stars. As an example, the well-studied CTTS TW Hya has a $\sim 4100$ K photosphere measured from high resolution optical spectra \citep[e.g.][]{yang05}.  However, the near-IR spectrum is consistent with $\sim M2$, when compared with main sequence dwarf stars \citep{vacca11}.  \citet{debes13} reconciled this discrepancy by invoking spots, although the spectral contribution from accretion and disk continuum emission was uncertain.  \citet{mcclure13} found that the TW Hya spectrum was well reproduced with near-IR spectra of stars with similar ($\sim$K7) optical spectral type, as long as the comparison spectra were young.   \citet{herczeg14} found an intermediate spectral type for TW Hya and other TTSs by focusing on TiO bands at redder wavelengths than previous spectra, and also by using young stars as photospheric templates.  {\bf Similar discrepancies have been found between optical and near-IR spectral types \citep[e.g.][]{bouvier92,bary14,cottaar14} and in comparisons of spectral types measurement based on TiO bands and those based on high resolution blue spectra \citep[see discussion in][]{herczeg14}.}

{bf The mismatches between near-IR spectra of dwarf stars and young stars (along with some of the color discrepancies, see Section 5.1) is also likely explained by the prevalence and importance of cool spots in low mass stars.  For classical T Tauri stars, the veiling at $\sim 1$ $\mu$m is usually larger than can be explained by either most accretion models\footnote{\citet{ingleby13} invoked multiple accretion flows with a range of densities to explain this red excess.} or warm disk emission \citep[e.g.][]{basri90,cieza05,fischer11,mcclure13}.  Similarly, \citet{cottaar14} found that most cool stars in the Pleiades are veiled in the H-band, with no possibility for any disk or accretion emission.  Differences in spot temperatures and coverages likely contribute to and may explain entirely this excess veiling.}

%[though see][][for a fit with multi-component accretion flow]{

%\subsection{Possibility of third temperature or multiple temperatures}

@@

%Assuming that the hotter component is 4100 K, the H$_2$O opacity is only sufficient if $T_{spot}<3000$ K, with a high spot coverage.

@@

%Introduce

@@

\section{Conclusions}
{\bf rearranged a bit}

This paper establishes that the optical and near-IR spectra of the weak-lined T Tauri star LkCa 4 are produced by multiple temperature components.   When the temperature distribution is simplified to two temperatures, the photospheric model that best fits the spectrum includes a hot (4100 K) photospheric region that covers 20\% of the stellar surface and a cool (2700--3000 K) photospheric that covers 80\% of the stellar surface.  The previously-established $V$-band brightness variability of $0.8$ mag is caused by the rotational modulation of the covering fraction of the hot photospheric component.  This spectral modeling leads to an improved effective temperature measurement of LkCa 4.  {\bf For LkCa 4 and other young low-mass stars, the standard approach of inferring masses and ages of young stars is inappropriate and will lead to large uncertainties, both because effective temperatures are systematically too hot and because most models of pre-main sequence evolution neglect spots}.

>>>>

Ages and masses of young stars are often estimated by comparing their observed luminosities and effective temperatures to pre-main sequence stellar evolutionary model tracks, but nonstandard physics like magnetic fields and starspots complicate this process. In order to understand their influence, we study in detail the heavily-spotted weak-lined \emph{T-Tauri} star LkCa 4 by searching for spectral signatures of radiation originating from the starspot or starspot groups.  We introduce a new methodology for constraining both the starspot filling factor and the typical spot temperature by fitting two-temperature stellar atmosphere models constructed from \PHOENIX\ synthetic spectra to a high-resolution near-IR IGRINS spectrum.  Clearly discernable spectral features arise from both a hot photospheric component $\teffa\sim4100$ K and to a cool component $\teffb\sim2700-3000$ K, with a best-fit filling factor of $\sim80$\%  for the cool photosphere at the time of the observation.  These results are supported by comparisons of this mix of hot and cool photospheric temperatures by analyses of the spectral energy distribution, rotational modulation of colors and of TiO band strengths, and low-resolution optical/near-IR spectroscopy.  LkCa 4 has either a warm photosphere mostly covered in cool spots or is a very cool star with a small hot spot.  The revised effective temperature and luminosity make LkCa 4 appear much younger and lower mass when compared to previous estimates and naively using the same model tracks.  Biases from starspots likely afflict most fully convective young stars and contribute to uncertainties in open cluster ages inferred from low-mass stars and in large age spreads of those clusters. We also discuss the detailed influence of starspots on the IGRINS spectrum, which in some regions acts as a relatively featureless ``veiling'' continuum owing to the large rotational broadening ($\vsini\sim28$ km~s$^{-1}$) and heavy line-blanketing in cool star spectra, and anti-correlations expected in the velocity jitter of the warm and cool components.

@@

The causes of age uncertainties and of large luminosity spreads in young clusters are controversial \citep[see reviews by][]{preibisch12,soderblom14}.  Absolute ages are usually inferred from comparisons between observed stellar properties and pre-main sequence evolutionary models, which are sensitive to stellar birthlines and the input physics that controls the contraction and internal heating.  Within clusters, any real differences in ages between stars may be masked by observational uncertainties in temperature and luminosity measurements \citep[e.g.][]{hartmann01,slesnick08}.

@@

In this work, we measure the starspot properties of LkCa 4 with four complementary techniques.  Section \ref{sec:obs} describes the full spectroscopic and photometric dataset and 
quantifies the optical variability of LkCa 4 over the last 31 years based on all available photometric monitoring; a collection of spectral observations at various phases of variability is introduced.  Section 3 and Appendix \ref{methods-details} describe a spectral inference technique aimed at generalizing line depth ratio analysis, which is applied to a high resolution $H$ and $K$ near-IR spectrum of LkCa 4 in Section \ref{sec:two_tempIGRINS}.  Section \ref{sec:GJHsection4} shows results from SED fitting, polychromatic photometric monitoring, and optical TiO line-depth ratio fitting.  All lines of evidence are ultimately combined to build a consistent picture of the spectral and temporal evolution of LkCa 4.  Section 5 describes how the two-temperature fit affects the placement of LkCa 4 on the HR diagram and what can be understood about pre-main sequence stellar evolution from this exemplar.  Finally, Section 6 describes how our results should be interpreted, the implications for analyses of spots in high-resolution spectra, and methodological limitations.

@@

Starspots possess a spectrum distinct from the ambient photosphere.  Approaches like ZDI and line-bisector analysis \citep[\emph{e.g.}][]{prato08, donati14} target the modulation of stellar photospheric lines as starspots enter and exit the observable stellar disk.  In contrast, the technique described in this Section measures the spectrum that emerges from both the hot and the cool surfaces of the photosphere.

@@

@@

Overplotting forward-modeled spectra with the observed IGRINS spectra yielded insights on why some spectral orders perform better than others in assessing stellar properties.  Orders with extremely poor telluric correction residuals, large spectral line outliers, and uncorrected H line residuals from A0V standard division, are all excluded from our final stellar parameter compilation.  Several orders in $K-$band were rejected due to poor spectral fits influenced by deep metal lines.  Metal lines can be biased by Zeeman broadening, which could explain poorer fit quality with wavelength \citep{johnskrull07,deen13}.  Additionally, some orders without these conspicuous faults, but simply possessing mediocre fits, or relatively uninformative fits, were discarded.  The remaining orders are relatively devoid of spectral line outliers and include the most information rich portion of the spectrum.

@@

@@

The synthetic models match the full spectrum reasonably well.  When scaled separately to the red-optical and near-IR spectra, the synthetic models match even better (bottom panels of Figure \ref{fig:lores}), while single temperature photospheres fail to reproduce large spectral features.   The $3900$ K component fits the TiO bands at $<7400$ \AA\ but fails to reproduce molecular bands at longer wavelengths.  The $3500$ K component suffers from the opposite problem, yielding molecular bands at short wavelengths that are too deep.  The two temperaturephotosphere with the parameters from the best-fit calculated in \S 3 reproduces the TiO bands, the hump in the $H$-band, and the jump in flux at the long-wavelength end of the $J$-band.  This latter feature is caused by H$_2$O opacity in H$_2$O bands and is only seen when the fraction of emission from the cool component is high (Figure \ref{fig:h2ojump}).

@@

The TiO bands trace the cool component.  When the hotter component has a larger filling fraction, the TiO bands are veiled by the hotter component and should therefore be shallower. As expected, 
the TiO indices all correlate strongly with the optical brightness.  The TiO-7140 band depths are accurately reproduced by varying the two-temperature fit with rotation.  Other bands have small offsets, likely caused by either contributions from intermediate temperatures or by small errors in the synthetic spectra.  In the spectral type scheme derived by \citet{herczeg14}, the change in TiO-7140 index corresponds to a range from M1.5 (3640 K) to M2.2 (3530 K).

@@

An accurate extinction estimate from the full SED is challenging because of a strong dependence on the  distribution of temperatures in the photosphere, which in our work is simplified to a two temperature fit.  However, isolating specific colors constrains the possible range of extinctions.  Based on the following analysis, an $A_V=0.3$ mag is adopted throughout this paper, although some uncertainty is introduced by confusion in the choice of template colors.

@@

Our results tentatively support the lower extinctions obtained at optical wavelengths, although the template colors used to calculate extinction are uncertain and will need revision.  Differences between main sequence and pre-main sequence temperature scales and colors in the \citet{pecaut13} tables are introduced by a combination of spots, in addition to any color differences caused by gravity differences.
Uncertainty in colors may be solved with self-consistent comparisons to young stars with similar spectral types, if spot coverage is also similar.  Discrepancies in colors, spectral type/effective temperatures, and extinctions for LkCa 4, and perhaps more broadly for TTSs, are dominated by ignoring spots and not by spectral type-effective temperature scales.  Fundamental problems in molecular opacities and photospheric models are likely important secondary affects.

@@

@@

The Pleiades serves as an older \citep[$\sim125$ Myr][]{stauffer98} sample of young stars, many of which have starspots that cover large fractions of the stellar surface.  For example, anomalous colors of K dwarfs in the Pleiades can be explained by $\sim$50\% fill factors of spots \citep{stauffer03}, while LAMOST spectra of the TiO band indicates that $\sim 30$\% spot coverages may be common for 3500--5000 K stars \citep{fang2016}.  Recent photometric monitoring of the Pleiades indicates that the anomalous colors scale with rotation rate, strengthing the case for magnetic-field induced starspots as the cause of the anomaly \citep{covey16}.  However, the \citet{covey16} Pleiades rotation rates do not scale with the $\Delta V$ photometric amplitude at the precision of the PTF data, suggesting that while the overall starspot coverage increases with rotation rate, the longitudinally asymmetric component does not.  LkCa 4 represents an example of such an architecture, with more than  50\% of the visible stellar surface covered in spots at all observational phases. The longitudinally symmetric component evades photometric modulation detection, but contributes to otherwise anomalous color offsets.

Spots are also likely important in estimating the properties of generic low-mass young stars. As an example, the well-studied CTTS TW Hya has a $\sim 4100$ K photosphere measured from high resolution optical spectra \citep[e.g.][]{yang05}.  However, the near-IR spectrum is consistent with $\sim M2$, when compared with main sequence dwarf stars \citep{vacca11}.  \citet{debes13} reconciled this discrepancy by invoking spots, although the spectral contribution from accretion and disk continuum emission was uncertain.  \citet{mcclure13} found that the TW Hya spectrum was well reproduced with near-IR spectra of stars with similar ($\sim$K7) optical spectral type, as long as the comparison spectra were young.   \citet{herczeg14} found an intermediate spectral type for TW Hya and other TTSs by focusing on TiO bands at redder wavelengths than previous spectra, and also by using young stars as photospheric templates.  Similar discrepancies have been found between optical and near-IR spectral types \citep[e.g.][]{bouvier92,bary14,cottaar14} and in comparisons of spectral types measurement based on TiO bands and those based on high resolution blue spectra \citep[see discussion in][]{herczeg14}.

The mismatches between near-IR spectra of dwarf stars and young stars (along with some of the color discrepancies, see Section 5.1) is also likely explained by the prevalence and importance of cool spots in low mass stars.  For classical T Tauri stars, the veiling at $\sim 1$ $\mu$m is usually larger than can be explained by either most accretion models\footnote{\citet{ingleby13} invoked multiple accretion flows with a range of densities to explain this red excess.} or warm disk emission \citep[e.g.][]{basri90,cieza05,fischer11,mcclure13}.  Similarly, \citet{cottaar14} found that most cool stars in the Pleiades are veiled in the H-band, with no possibility for any disk or accretion emission.  Differences in spot temperatures and coverages likely contribute to and may explain entirely this excess veiling.

@@

@@

@@

\section{Conclusions}

This paper establishes that the optical and near-IR spectra of the weak-lined T Tauri star LkCa 4 are produced by multiple temperature components.   When the temperature distribution is simplified to two temperatures, the photospheric model that best fits the spectrum includes a hot (4100 K) photospheric region that covers 20\% of the stellar surface and a cool (2700--3000 K) photospheric that covers 80\% of the stellar surface.  The previously-established $V$-band brightness variability of $0.8$ mag is caused by the rotational modulation of the covering fraction of the hot photospheric component.  This spectral modeling leads to an improved effective temperature measurement of LkCa 4. For LkCa 4 and other young low-mass stars, the standard approach of inferring masses and ages of young stars is inappropriate and will lead to large uncertainties, both because effective temperatures are systematically too hot and because most models of pre-main sequence evolution neglect spots.
